---
title: "Kalman"
---

Sure, here is an example of how you might use the Kalman filter to predict customer churn in a banking context using R and the tidyverse package. Note that this is a simplified illustration and assumes you have a dataset with relevant variables.

### Sample Data

Let's assume we have a dataset `customer_data` with the following columns: - `customer_id`: Unique identifier for each customer. - `transaction_freq`: Frequency of transactions in the current month. - `avg_balance`: Average account balance in the current month. - `cust_service_calls`: Number of customer service interactions in the current month. - `churn`: Whether the customer churned (1) or not (0).

### Sample Code

``` r
# Load necessary libraries
library(tidyverse)
library(dplyr)
library(KFAS)

# Generate sample data
set.seed(123)
customer_data <- tibble(
  customer_id = 1:100,
  transaction_freq = rnorm(100, mean = 10, sd = 2),
  avg_balance = rnorm(100, mean = 5000, sd = 1000),
  cust_service_calls = rpois(100, lambda = 2),
  churn = sample(c(0, 1), 100, replace = TRUE)
)

# Define the state space model
# For simplicity, we'll use a local level model for each state variable
# Assume normal distribution for observations

# Build a state space model
build_model <- function(data) {
  ssmodel <- SSModel(
    H = matrix(NA, 3, 3),
    Q = diag(NA, 3),
    Z = diag(3),
    T = diag(3),
    R = diag(3),
    a1 = c(mean(data$transaction_freq), mean(data$avg_balance), mean(data$cust_service_calls)),
    P1 = diag(NA, 3)
  )
  
  ssmodel$y <- t(as.matrix(data %>% select(transaction_freq, avg_balance, cust_service_calls)))
  return(ssmodel)
}

# Fit the model
fit_model <- function(ssmodel) {
  fit <- fitSSM(ssmodel, inits = rep(0, 6), method = "BFGS")
  return(fit$model)
}

# Kalman filter application
apply_kalman_filter <- function(model) {
  kf <- KFS(model, filtering = "mean", smoothing = "mean")
  return(kf)
}

# Predict churn probabilities based on filtered states
predict_churn <- function(kf, threshold = 0.5) {
  smoothed_states <- kf$a
  churn_prob <- 1 / (1 + exp(-rowMeans(smoothed_states))) # Logistic function for churn probability
  churn_pred <- ifelse(churn_prob > threshold, 1, 0)
  return(churn_pred)
}

# Build, fit, and apply the Kalman filter model
ssmodel <- build_model(customer_data)
fitted_model <- fit_model(ssmodel)
kf_results <- apply_kalman_filter(fitted_model)

# Predict churn
customer_data <- customer_data %>%
  mutate(predicted_churn = predict_churn(kf_results))

# Display results
print(customer_data)

# Evaluate the model
confusion_matrix <- table(customer_data$churn, customer_data$predicted_churn)
print(confusion_matrix)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", accuracy))
```

### Explanation:

1.  **Data Generation**: Simulate customer data for the example.
2.  **State Space Model Definition**: Define a simple state space model where each state variable is modeled independently.
3.  **Model Fitting**: Fit the state space model using the Kalman filter.
4.  **Filtering and Smoothing**: Apply the Kalman filter and smoother to estimate the hidden states.
5.  **Churn Prediction**: Use the smoothed states to predict churn probabilities, applying a logistic function.
6.  **Evaluation**: Compare the predicted churn with the actual churn to evaluate model performance.

This code provides a simplified illustration. In a real-world application, you would need to preprocess the data, handle missing values, and potentially use more complex state space models to capture the dynamics of customer behavior more accurately.

::: callout-note
```         
Please cite KFAS in publications by using: 

Jouni Helske (2017). KFAS: Exponential Family State Space Models in R. Journal of Statistical Software, 78(10), 1-39. doi:10.18637/jss.v078.i10.
```
:::

```{r}
#| eval: false
library(tidyverse)
library(dplyr)
library(KFAS)

# Generate sample data
set.seed(123)
customer_data <- tibble(
  customer_id = 1:100,
  transaction_freq = rnorm(100, mean = 10, sd = 2),
  avg_balance = rnorm(100, mean = 5000, sd = 1000),
  cust_service_calls = rpois(100, lambda = 2),
  churn = sample(c(0, 1), 100, replace = TRUE)
)

customer_data
```

```{r}
#| eval: false
# Define the state space model
# For simplicity, we'll use a local level model for each state variable
# Assume normal distribution for observations

# Build a state space model
# build_model <- function(data) {
#   ssmodel <- KFAS::SSModel(
#     H = matrix(NA, 3, 3),
#     Q = diag(NA, 3),
#     Z = diag(3),
#     T = diag(3),
#     R = diag(3),
#     a1 = c(mean(data$transaction_freq), mean(data$avg_balance), mean(data$cust_service_calls)),
#     P1 = diag(NA, 3)
#   )
#   
#   ssmodel$y <- t(as.matrix(data %>% select(transaction_freq, avg_balance, cust_service_calls)))
#   return(ssmodel)
# }

build_model <- function(data) {
  # Define the state space model using a formula
  model <- SSModel(
    transaction_freq + avg_balance + cust_service_calls ~ 
      SSMtrend(1, Q = list(NA)),  # Local level model for each variable
    H = matrix(NA, 3, 3),  # Observation noise covariance matrix
    data = data
  )
  
  # Assign observed data to the model
  model$y <- t(as.matrix(data %>% select(transaction_freq, avg_balance, cust_service_calls)))
  
  return(model)
}

# Fit the model
fit_model <- function(ssmodel) {
  fit <- fitSSM(ssmodel, inits = rep(0, 6), method = "BFGS")
  return(fit$model)
}

# Kalman filter application
apply_kalman_filter <- function(model) {
  kf <- KFS(model, filtering = "mean", smoothing = "mean")
  return(kf)
}

# Predict churn probabilities based on filtered states
predict_churn <- function(kf, threshold = 0.5) {
  smoothed_states <- kf$a
  churn_prob <- 1 / (1 + exp(-rowMeans(smoothed_states))) # Logistic function for churn probability
  churn_pred <- ifelse(churn_prob > threshold, 1, 0)
  return(churn_pred)
}

```

```{r}
#| eval: false
# Build, fit, and apply the Kalman filter model
ssmodel <- build_model(customer_data)
fitted_model <- fit_model(ssmodel)
kf_results <- apply_kalman_filter(fitted_model)

# Predict churn
customer_data <- customer_data %>%
  mutate(predicted_churn = predict_churn(kf_results))

```
