---
title: "Bayes"
---

{{< include 00-setup.qmd >}}

```{r}
#| message: false
#| warning: false
#| echo: false
# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(
  tidyverse, broom, rsample, ggdag, causaldata, halfmoon, ggokabeito
  , magrittr, ggplot2, estimatr, Formula, gt, gtExtras,
  brms, tidybayesm, bayesplot, ggthemes, patchwork)

# set the default theme for plotting
theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))
```

## Basic model

### Coin flip

We express the likelihood for our coin toss example as

$$y_{i} \sim \operatorname{Bernoulli}(\theta)$$

Our prior will be

$$\theta \sim \operatorname{Beta}(\alpha, \beta)$$

```{r}
#|echo: true
my_data <- readr::read_csv("data/z15N50.csv", show_col_types = FALSE)
summary(my_data)

my_data |> 
  dplyr::mutate(y = y |> as.character()) |> 
  ggplot(aes(x = y)) +
  geom_bar() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()
```

```{r}
#| message: false
#| output: false

fit8.1 <-
  brms::brm(data = my_data, 
      family = brms::bernoulli(link = identity),
      formula = y ~ 1,
      brms::prior(beta(2, 2), class = Intercept, lb = 0, ub = 1),
      iter = 500 + 3334, warmup = 500, chains = 3,
      seed = 8,
      file = "fits/fit08.01")
```

```{r}
#| message: false
plot(fit8.1)
```

```{r}
#| message: false
#| output: false
draws <- brms::as_draws_df(fit8.1) 
draws

```

```{r}
draws |> 
  dplyr::mutate(chain = .chain) |> 
  bayesplot::mcmc_dens_overlay(pars = vars(b_Intercept)) 

```

```{r}
draws |> 
  dplyr::mutate(chain = .chain) |> 
  bayesplot::mcmc_acf(pars = vars(b_Intercept), lags = 35) +
  theme_minimal()
```

```{r}
brms::rhat(fit8.1)["b_Intercept"]
brms::neff_ratio(fit8.1)["b_Intercept"]
```

```{r}
print(fit8.1)
```

```{r}
brms::posterior_summary(fit8.1, robust = T)
```

```{r}
brms::posterior_summary(fit8.1, probs = c(.025, .25, .75, .975))

```

```{r}
bayesplot::mcmc_areas(
  draws, 
  pars = vars(b_Intercept),
  prob = 0.5,
  prob_outer = 0.95,
  point_est = "mean"
) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Theta via bayesplot::mcmc_areas()",
       x = expression(theta)) +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

```{r}
library(tidybayes)

draws |> 
  ggplot(aes(x = b_Intercept, y = 0)) +
  tidybayes::stat_halfeye(point_interval = tidybayes::mode_hdi, .width = c(.95, .5)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = expression(theta*" via tidybayes::stat_halfeye()"),
       x = expression(theta)) +
  theme_minimal()

draws |> 
  ggplot(aes(x = b_Intercept, y = 0)) +
  tidybayes::stat_histinterval(point_interval = tidybayes::mode_hdi, .width = c(.95, .5)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = expression(theta*" via tidybayes::stat_histinterval()"),
       x = expression(theta)) +
  theme_minimal()
```

```{r}
dat <- readr::read_csv("data/z6N8z2N7.csv", show_col_types = FALSE) |> 
  dplyr::mutate(
    s = 
      dplyr::case_when(s == "Reginald" ~ "Windsor", TRUE ~ "London")
  )
```

```{r}
dat |> 
  dplyr::mutate(y = y |> as.character()) |> 
  ggplot(aes(x = y, fill = s)) +
  geom_bar(show.legend = F) +
  ggthemes::scale_fill_colorblind() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  facet_wrap(~ s)
```

```{r}
#| output: false
#| echo: true
#| label: fit8.2
fit8.2 <-
  brms::brm(data = dat, 
      family = brms::bernoulli(identity),
      y ~ 0 + s,
      brms::prior(beta(2, 2), class = b, lb = 0, ub = 1),
      iter = 2000, warmup = 500, cores = 4, chains = 4,
      seed = 8,
      file = "fits/fit08.02")
```

```{r}
plot(fit8.2, widths = c(2, 3))
```

```{r}
summary(fit8.2)
```

```{r}
pairs(fit8.2,
      off_diag_args = list(size = 1/3, alpha = 1/3))
```

```{r}
draws <- brms::as_draws_df(fit8.2)

draws <-
  draws |> 
  dplyr::rename(theta_Windsor = b_sWindsor,
         theta_London  = b_sLondon) |> 
  dplyr::mutate(`theta_Windsor - theta_London` = theta_Windsor - theta_London)

dplyr::glimpse(draws |> dplyr::slice_head(n=6))
```

```{r}
#| warning: false
#| echo: true
long_draws <-
  draws |> 
  dplyr::select(starts_with("theta")) |> 
  tidyr::pivot_longer(everything()) |> 
  dplyr::mutate(name = factor(name, levels = c("theta_Windsor", "theta_London", "theta_Windsor - theta_London"))) 
  
long_draws |> 
  ggplot(aes(x = value, y = 0, fill = name)) +
  stat_histinterval(point_interval = mode_hdi, .width = .95,
                    slab_color = "white", outline_bars = T,
                    normalize = "panels") +
  scale_fill_manual(values = ggthemes::colorblind_pal()(8)[2:4], breaks = NULL) +
  scale_y_continuous(NULL, breaks = NULL) +
  theme_minimal() +
  facet_wrap(~ name, scales = "free")
```

```{r}
long_draws |> 
  dplyr::group_by(name) |> 
  tidybayes::mode_hdi()
```

```{r}
#| output: false
#| warning: false
#| echo: true
#| label: fit8.3

fit8.3 <-
  brms::brm(data = dat, 
      family = brms::bernoulli(identity),
      y ~ 0 + s,
      prior =
        c(brms::prior(beta(2, 2), class = b, coef = sWindsor),
          brms::prior(beta(2, 2), class = b, coef = sLondon),
          # this just sets the lower and upper bounds
          brms::prior(beta(2, 2), class = b, lb = 0, ub = 1)),
      iter = 2000, warmup = 500, cores = 4, chains = 4,
      sample_prior = "only",
      seed = 8,
      file = "fits/fit08.03")
```

```{r}
draws <- brms::as_draws_df(fit8.3) |> 
  dplyr::select(starts_with("b_"))

head(draws)
```

```{r}
dat |> 
  dplyr::group_by(s) |> 
  dplyr::summarise(z = sum(y),
            N = dplyr::n()) |> 
  dplyr::mutate(`z/N` = z / N)

levels <- c("theta_Windsor", "theta_London", "theta_Windsor - theta_London")

d_line <-
  tibble::tibble(value = c(.75, .286, .75 - .286),
         name  =  factor(c("theta_Windsor", "theta_London", "theta_Windsor - theta_London"), 
                         levels = levels))
```

```{r}
draws |> 
  dplyr::rename(theta_Windsor = b_sWindsor,
         theta_London  = b_sLondon) |> 
  dplyr::mutate("theta_Windsor - theta_London" = theta_Windsor - theta_London) |> 
  tidyr::pivot_longer(contains("theta")) |> 
  dplyr::mutate(name = factor(name, levels = levels)) |>
  
  ggplot(aes(x = value, y = 0)) +
  stat_histinterval(point_interval = tidybayes::mode_hdi, .width = .95,
                    fill = ggthemes::colorblind_pal()(8)[5], normalize = "panels") +
  geom_vline(data = d_line, 
             aes(xintercept = value), 
             linetype = 2) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(subtitle = expression("The dashed vertical lines mark off "*italic(z[s])/italic(N[s]))) +
  cowplot::theme_cowplot() +
  facet_wrap(~ name, scales = "free")
```

```{r}
draws |> 
  dplyr::rename(theta_Windsor = b_sWindsor,
         theta_London  = b_sLondon) |> 
  
  ggplot(aes(x = theta_Windsor, y = theta_London)) +
  geom_point(alpha = 1/4, color = ggthemes::colorblind_pal()(8)[6]) +
  coord_equal() +
  cowplot::theme_minimal_grid()
```

```{r}
draws |> 
  dplyr::rename(theta_Windsor = b_sWindsor,
         theta_London  = b_sLondon) |> 
  
  ggplot(aes(x = theta_Windsor, y = theta_London)) +
  stat_density_2d(aes(fill = stat(density)), 
                  geom = "raster", contour = F) +
  scale_fill_viridis_c(option = "B", breaks = NULL) +
  scale_x_continuous(expression(theta[1]), 
                     expand = c(0, 0), limits = c(0, 1),
                     breaks = 0:4 / 4, labels = c("0", ".25", ".5", ".75", "1")) +
  scale_y_continuous(expression(theta[2]), 
                     expand = c(0, 0), limits = c(0, 1),
                     breaks = 0:4 / 4, labels = c("0", ".25", ".5", ".75", "1")) +
  coord_equal() +
  cowplot::theme_minimal_grid()
```

## Hierarchical model

### Coin flip

## A single coin from a single mint

Recall from the last chapter that our likelihood is the Bernoulli distribution,

$$y_i \sim \operatorname{Bernoulli}(\theta).$$

We'll use the beta density for our prior distribution for $\theta$,

$$\theta \sim \operatorname{Beta}(\alpha, \beta).$$

And we can re-express $\alpha$ and $\beta$ in terms of the mode $\omega$ and concentration $\kappa$, such that

$$\alpha = \omega(\kappa - 2) + 1 \;\;\; \textrm{and} \;\;\; \beta = (1 - \omega)(\kappa - 2) + 1.$$

As a consequence, we can re-express $\theta$ as

$$\theta \sim \operatorname{Beta}(\omega(\kappa - 2) + 1, (1 - \omega)(\kappa - 2) + 1).$$

The value of $\kappa$ governs how near $\theta$ is to $\omega$, with larger values of $\kappa$ generating values of $\theta$ more concentrated near $\omega$.

Using $s$ for shape and $r$ for rate, Kruschke's Equations 9.7 and 9.8 are as follows:

$$
s = \frac{\mu^2}{\sigma^2} \;\;\; \text{and} \;\;\; r = \frac{\mu}{\sigma^2} \;\;\; \text{for mean} \;\;\; \mu > 0 \\
s = 1 + \omega r \;\;\; \text{where} \;\;\; r = \frac{\omega + \sqrt{\omega^2 + 4\sigma^2}}{2\sigma^2} \;\;\; \text{for mode} \;\;\; \omega > 0.
$$

```{r}
gamma_s_and_r_from_mean_sd <- function(mean, sd) {
  if (mean <= 0) stop("mean must be > 0")
  if (sd   <= 0) stop("sd must be > 0")
  shape <- mean^2 / sd^2
  rate  <- mean   / sd^2
  return(list(shape = shape, rate = rate))
}

gamma_s_and_r_from_mode_sd <- function(mode, sd) {
  if (mode <= 0) stop("mode must be > 0")
  if (sd   <= 0) stop("sd must be > 0")
  rate  <- (mode + sqrt(mode^2 + 4 * sd^2)) / (2 * sd^2)
  shape <- 1 + mode * rate
  return(list(shape = shape, rate = rate))
}
```

```{r}
gamma_s_and_r_from_mean_sd(mean = 10, sd = 100)
gamma_s_and_r_from_mode_sd(mode = 10, sd = 100)
```

```{r}
dat <- readr::read_csv("data/TherapeuticTouchData.csv", show_col_types = FALSE)

dplyr::glimpse(dat)
```

```{r}
dat |> 
  dplyr::mutate(y = y |> as.character()) |> 
  
  ggplot(aes(y = y)) +
  geom_bar(aes(fill = after_stat(count))) +
  scale_fill_viridis_c(option = "A", end = .7, breaks = NULL) +
  scale_x_continuous(breaks = 0:4 * 2, expand = c(0, NA), limits = c(0, 9)) +
  cowplot::theme_minimal_vgrid() +
  cowplot::panel_border() +
  facet_wrap(~ s, ncol = 7)
```

```{r}
#| warning: false
#| code-fold: true
a_purple <- viridis::viridis_pal(option = "A")(9)[4]
dat |> 
  dplyr::group_by(s) |> 
  dplyr::summarize(mean = mean(y)) |>
  
  ggplot(aes(x = mean)) +
  geom_histogram(color = "white", fill = a_purple,
                 linewidth = .2, binwidth = .1) +
  scale_x_continuous("Proportion Correct", limits = c(0, 1)) +
  scale_y_continuous("# Practitioners", expand = c(0, NA)) +
  cowplot::theme_minimal_hgrid()
```

```{r}
#| warning: false
#| code-fold: true
#| message: false
#| label: fit9.1
fit9.1 <-
  brms::brm(data = dat,
      family = brms::bernoulli(link = logit),
      y ~ 1 + (1 | s),
      prior = c(brms::prior(normal(0, 1.5), class = Intercept),
                brms::prior(normal(0, 1), class = sd)),
      iter = 20000, warmup = 1000, thin = 10, chains = 4, cores = 4,
      seed = 9,
      file = "fits/fit09.01")
```

```{r}
plot(fit9.1, widths = c(2, 3))
```

```{r}
draws <- brms::as_draws_df(fit9.1)
```

```{r}
draws |> 
  dplyr::mutate(chain = .chain) |> 
  bayesplot::mcmc_acf(pars = vars(b_Intercept, sd_s__Intercept), lags = 10) +
  cowplot::theme_cowplot()
```

```{r}
bayesplot::neff_ratio(fit9.1) |> 
  bayesplot::mcmc_neff() +
  cowplot::theme_cowplot(font_size = 12)
```

The $N_\textit{eff}/N$ ratio values for our model parameters were excellent. Here's a numeric summary of the model.

```{r}
print(fit9.1)
```

We'll need `brms::inv_logit_scaled()` to convert the model parameters to predict $\theta$ rather than $\operatorname{logit}(\theta)$. After the conversions, we'll be ready to make the histograms in the lower portion of Figure 9.10.

```{r}
# wrangle
draws_small <-
  draws |> 
  # convert the linear model parameters to the probability space with `inv_logit_scaled()`
  dplyr::mutate(`theta[1]`  = (b_Intercept + `r_s[S01,Intercept]`) |> brms::inv_logit_scaled(),
         `theta[14]` = (b_Intercept + `r_s[S14,Intercept]`) |> brms::inv_logit_scaled(),
         `theta[28]` = (b_Intercept + `r_s[S28,Intercept]`) |> brms::inv_logit_scaled()) |> 
  # make the difference distributions
  dplyr::mutate(`theta[1] - theta[14]`  = `theta[1]`  - `theta[14]`,
         `theta[1] - theta[28]`  = `theta[1]`  - `theta[28]`,
         `theta[14] - theta[28]` = `theta[14]` - `theta[28]`) |> 
  dplyr::select(starts_with("theta"))

draws_small |> 
  tidyr::pivot_longer(everything()) |> 
  # this line is unnecessary, but will help order the plots 
  dplyr::mutate(name = factor(name, levels = c("theta[1]", "theta[14]", "theta[28]", 
                                        "theta[1] - theta[14]", "theta[1] - theta[28]", "theta[14] - theta[28]"))) |> 

  ggplot(aes(x = value, y = 0)) +
  stat_histinterval(point_interval = mode_hdi, .width = .95,
                    fill = a_purple, breaks = 40, normalize = "panels") +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab(NULL) +
  cowplot::theme_minimal_hgrid() +
  facet_wrap(~ name, scales = "free", ncol = 3)
```

```{r}
draws_small |> 
  tidyr::pivot_longer(everything()) |>
  dplyr::group_by(name) |> 
  tidybayes::mode_hdi(value) |> 
  dplyr::mutate_if(is.double, round, digits = 3)
```

```{r}
p1 <-
  draws_small |> 
  ggplot(aes(x = `theta[1]`, y = `theta[14]`)) +
  geom_abline(linetype = 2) +
  geom_point(color = a_purple, size = 1/8, alpha = 1/8)

p2 <-
  draws_small |> 
  ggplot(aes(x = `theta[1]`, y = `theta[28]`)) +
  geom_abline(linetype = 2) +
  geom_point(color = a_purple, size = 1/8, alpha = 1/8)

p3 <-
  draws_small |> 
  ggplot(aes(x = `theta[14]`, y = `theta[28]`)) +
  geom_abline(linetype = 2) +
  geom_point(color = a_purple, size = 1/8, alpha = 1/8)

(p1 + p2 + p3) &
  coord_cartesian(xlim = c(0, 1),
                  ylim = c(0, 1)) & 
  cowplot::theme_minimal_grid()
```

```{r}
#| warning: false
#| code-fold: true
#| message: false
# this part makes it easier to set the break points in `scale_x_continuous()` 
labels <-
  draws |> 
  dplyr::mutate(theta = b_Intercept |> inv_logit_scaled(), .keep = "none") |>
  tidybayes::mode_hdi() |> 
  tidyr::pivot_longer(theta:.upper) |> 
  dplyr::mutate(label = value |> round(3) |> as.character())
  
draws |> 
  mutate(theta = b_Intercept |> inv_logit_scaled()) |> 

  ggplot(aes(x = theta, y = 0)) +
  stat_histinterval(point_interval = mode_hdi, .width = .95,
                    fill = a_purple, breaks = 40) +
  scale_x_continuous(expression(theta), 
                     breaks = labels$value,
                     labels = labels$label) +  
  scale_y_continuous(NULL, breaks = NULL) +
  cowplot::theme_minimal_hgrid()
```

```{r}
# the tibble of the primary data
sigmas <-
  stats::coef(fit9.1, summary = F)$s |> 
  tibble::as_tibble() |> 
  dplyr::mutate(iter = 1:n()) |> 
  dplyr::group_by(iter) |> 
  tidyr::pivot_longer(-iter) |> 
  dplyr::mutate(theta = brms::inv_logit_scaled(value)) |> 
  dplyr::summarise(sd = sd(theta))

# this, again, is just to customize `scale_x_continuous()`
labels <-
  sigmas |> 
  tidybayes::mode_hdi(sd) |> 
  tidyr::pivot_longer(sd:.upper) |> 
  dplyr::mutate(label = value |> round(3) |> as.character())
  
# the plot
sigmas |> 
  ggplot(aes(x = sd, y = 0)) +
  stat_histinterval(point_interval = mode_hdi, .width = .95,
                    fill = a_purple, breaks = 40) +
  scale_x_continuous(expression(paste(sigma, " of ", theta, " in a probability metric")),
                     breaks = labels$value,
                     labels = labels$label) +  
  scale_y_continuous(NULL, breaks = NULL) +
  cowplot::theme_minimal_hgrid()
```

```{r}
#| warning: false
#| code-fold: true
#| message: false
color_scheme_set("purple")
bayesplot::bayesplot_theme_set(theme_default() + cowplot::theme_minimal_grid())

stats::coef(fit9.1, summary = F)$s |> 
  brms::inv_logit_scaled() |> 
  data.frame() |> 
  rename(`theta[1]`  = S01.Intercept, 
         `theta[14]` = S14.Intercept, 
         `theta[28]` = S28.Intercept) |> 
  dplyr::select(`theta[1]`, `theta[14]`, `theta[28]`) |> 
  
  bayesplot::mcmc_pairs(off_diag_args = list(size = 1/8, alpha = 1/8))
```

Did you see how we slipped in the `color_scheme_set()` and `bayesplot_theme_set()` lines at the top? Usually, the plots made with **bayesplot** are easy to modify with **ggplot2** syntax. Plots made with `mcmc_pairs()` function are one notable exception. On the back end, these made by combining multiple **ggplot** into a grid, a down-the-line result of which is they are difficult to modify. Happily, one can make some modifications beforehand by altering the global settings with the `color_scheme_set()` and `bayesplot_theme_set()` functions. You can learn more in the discussion on [issue #128](https://github.com/stan-dev/bayesplot/issues/128) on the **bayesplot** GitHub repo.
