---
title: "Movie budgets and revenues"
editor: visual
---

::: callout-important
This application exercise is a demo only. You do not have a corresponding repository for it and you're not expected to turn in anything for it.
:::

We will look at the relationship between budget and revenue for movies made in the United States in 1986 to 2020. The dataset is created based on data from the [Internet Movie Database (IMDB)](https://www.imdb.com/).

```{r load-packages}
#| warning: false
#| message: false

# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(
  tidyverse  # for data analysis and visualisation
  , magrittr # for piping data between operations
  , scales   # for pretty axis labels
  , DT       # for interactive table
)
```

## Data

The `movies` data set includes basic information about each movie including budget, genre, movie studio, director, etc. A full list of the variables may be found [here](https://github.com/danielgrijalva/movie-stats).

```{r read-data}
#| message: false
#| warning: false
movies <- 
  readr::read_csv(
    "https://raw.githubusercontent.com/danielgrijalva/movie-stats/master/movies.csv"
    , show_col_types = FALSE
    )
```

View the first 10 rows of data.

```{r}
movies
```

The \_\_\_ dataset has \_\_\_ observations and \_\_\_ variables.

## Analysis

### Gross over time

We begin by looking at how the average gross revenue (`gross`) has changed over time. Since we want to visualize the results, we will choose a few genres of interest for the analysis.

```{r genre-list}
genre_list <- c("Comedy", "Action", "Animation", "Horror")
```

Then, we will filter for these genres and visualize the average gross revenue over time.

```{r plot-rating}
#| echo: true
#| messages: false
#| warning: false
movies %>%
  dplyr::filter(genre %in% genre_list) %>% 
  dplyr::group_by(genre,year) %>%
  dplyr::summarise(avg_gross = mean(gross), .groups = "keep") %>%
  ggplot(mapping = aes(x = year, y = avg_gross, color= genre)) +
    geom_point() + 
    geom_line() +
    scale_color_viridis_d() +
    scale_y_continuous(labels = label_dollar()) +
    labs(
      x = "Year",
      y = "Average Gross Revenue (US Dollars)",
      color = "Genre",
      title = "Gross Revenue Over Time"
    )
```

The plot suggests ...

### Budget and gross

Next, let's see the relationship between a movie's budget and its gross revenue.

```{r plot-revenue}
#| echo: true
#| warning: false
movies %>%
  dplyr::filter(genre %in% genre_list, budget > 0) %>% 
  ggplot(mapping = aes(x=log(budget), y = log(gross), color=genre)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ genre) + 
  scale_color_viridis_d() +
  labs(
    x = "Log-transformed Budget",
    y = "Log-transformed Gross Revenue"
  )
```

## Exercises

1.  Suppose we fit a regression model for each genre that uses budget to predict gross revenue. What are the signs of the correlation between `budget` and `gross` and the slope in each regression equation?

2.  Suppose we fit the regression model from the previous question. Which genre would you expect to have the smallest residuals, on average (*residual = observed revenue - predicted revenue*)?

3.  In the remaining time, discuss the following: Notice in the graph above that `budget` and `gross` are log-transformed. Why are the log-transformed values of the variables displayed rather than the original values (in U.S. dollars)?

## Appendix

Below is a list of genres in the data set:

```{r}
movies %>% 
  distinct(genre) %>%
  arrange(genre) %>% 
  datatable()
```
