---
title: "Covariance/Variance Decomposition"
format: pdf
keep-tex: true
keep-md: true
header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}

# fig-pos: 'H'
#monofont: 'Source Code Pro'
monofontoptions: 
  - Scale=0.55
margin-left: 20mm
margin-right: 20mm
margin-top: 20mm
margin-bottom: 25mm
---

```{r}
#| include: false
#| echo: false

# use \centering with latex tables - post processing

# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(
viridis, ggplot2, patchwork, showtext, extrafont
)

sysfonts::font_add_google("Lato")
showtext::showtext_auto()
theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))
```

Here we show that the population covariance can be expressed as a difference equation.

For $\Delta_k a_t \equiv a_{t+k}-a_t$ (the difference) and $\bar{a}\equiv\frac{1}{T}\sum_{t=1}^T a_t$ (the mean)

Given sequences $\left\{a_t\right\}_{t=1}^T$ and $\left\{b_t\right\}_{t=1}^T$, we have

$$\frac{1}{T}\sum_{t=1}^T(a_t-\bar{a})(b_t-\bar{b})= \frac{1}{T^2}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k} \Delta_k a_t \Delta_k b_t$$

## Strategy

The population covariance can be written as follows:

\begin{align}
\frac{1}{T}\sum_{t=1}^{T}(a_{t}-\bar{a})(b_{t}-\bar{b}) & =\frac{1}{T}\sum_{t=1}^{T}\left(a_{t}b_{t}-\bar{a}b_{t}-a_{t}\bar{b}+\bar{a}\bar{b}\right)\\
 & =\left(\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}\right)-\bar{a}\bar{b}
\end{align}

so the proof strategy is to show that the difference equation can be reduced to the same form.

## Proof

Since $\Delta_k a_t \equiv a_{t+k}-a_t$ we can expand the difference equation as:

\begin{align}
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\Delta_{k}a_{t}\Delta_{k}b_{t} & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}-a_{t}\right)\left(b_{t+k}-b_{t}\right)\\
 & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}b_{t+k}+a_{t}b_{t}\right)\\
 & -\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t}b_{t+k}-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}
\end{align}

and we'll evaluate each of the 4 terms in turn.

### term 1:

Making a change of variable: $s=t+k$ and noting that when $t$ ranges from $1$ to $T − k$, $s$ ranges from $1+k$ to $T$, we have:

\begin{align}
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t+k} & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{s=k+1}^{T}a_{s}b_{s}\\
 & =\frac{1}{T^{2}}\sum_{s=2}^{T}\sum_{k=1}^{s-1}a_{s}b_{s}\\
 & =\frac{1}{T^{2}}\sum_{s=2}^{T}\left(s-1\right)a_{s}b_{s}
\end{align}

where in the second-last line we note that $a_s b_s$ doesn't depend on $k$ so we count each $a_s b_s$ exactly $s-1$ times for $s\in[2,T]$ (i.e. due to the inner sum). Note that this is

$$
\frac{1}{T^{2}}\left(a_{2}b_{2}+2a_{3}b_{3}+3a_{4}b_{4}+\cdots++\left(T-2\right)a_{T-1}b_{T-1}+\left(T-1\right)a_{T}b_{T}\right)
$$

### term 2:

The second term is

$$
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_t b_t
$$

and it can be decomposed directly to:

$$
\frac{1}{T^{2}}\left((T-1)a_{1}b_{1}+(T-2)a_{2}b_{2}+(T-3)a_{3}b_{3}+\cdots+a_{T-1}b_{T-1}\right)
$$


i.e. $\frac{1}{T^2}$ times the following sum of $T-1$ values:

\begin{align*}
+ & \left(a_{1}b_{1}+a_{2}b_{2}+\cdots+a_{T-3}b_{T-3}+a_{T-2}b_{T-2}+a_{T-1}b_{T-1}\right)\\
+ & \left(a_{1}b_{1}+a_{2}b_{2}+\cdots+a_{T-3}b_{T-3}+a_{T-2}b_{T-2}\right)\\
 & \qquad\qquad\qquad\vdots\\
+ & \left(a_{1}b_{1}+a_{2}b_{2}\right)\\
+ & \left(a_{1}b_{1}\right)
\end{align*}

So combining the first and second terms we get $\frac{T-1}{T^2}\sum_{t=1}^{T}a_t b_t$, and we're on our way towards showing that

$$
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\Delta_{k}a_{t}\Delta_{k}b_{t} = \left(\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}\right)-\bar{a}\bar{b}
$$

### term 3:

The third term is $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}$ and again we will use the change of variables $s=t+k$.

Before changing variables, the outer sum goes from $k=1$ to $k=T-1$ and the inner sum goes from $t=1$ to $t=T-k$. The $(k,t)$ indices used in $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}$ are shown in figure (A) below (for $T=10$), along with the $s=t+k$ values.

With the change of variable, the outer sum goes from $s=t+k=2$ to $s=t+k=T$ and the inner sum goes from $t=1$ to $t=s-1$. The $(s,t)$ indices used to sum $-\frac{1}{T^{2}}\sum_{s=2}^{T}\sum_{t=1}^{s-1}a_{s}b_{t}$ are shown in figure (B) below (for $T=10$).

The first index ($s$) is along the horizontal axis, while the second index ($t$) is along the vertical axis.

```{r}
#| label: calc
#| echo: true
#| fig-align: center

T <- 10
res <- 1:(T-1) |> # outer sum
  purrr::map( 
    (\(k){ 1:(T-k) |> # inner sum
        purrr::map(\(t)data.frame("k"=k,"t"=t, "s"=t+k)) }
    ) 
  ) |> dplyr::bind_rows()
```

```{r}
#| label: first grid plot
#| echo: false
#| fig-align: center
p1 <- res |> 
  ggplot(aes(x=k,y=t, label = as.character(s))) + geom_label(color = "#3B528B") +
  #geom_point(color = "#3B528B") +
  scale_x_continuous(breaks = 1:T) +
  scale_y_continuous(
    "t"
    , breaks = 1:T
    , labels = 1:T |> as.character()
  ) +
  theme_minimal()  +
  theme(axis.title.y.left = element_text(angle = 0, vjust = 0.5)) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "s values given k and t", subtitle="T=10")+ coord_fixed()

p2 <- res |> 
  ggplot(aes(x=s,y=t+1))+
  geom_point(shape=0, color = "#21908C") +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(name = "s", breaks = 1:T) +
  scale_y_continuous(
    sec.axis = ggplot2::sec_axis(transform = ~.-1, breaks = 1:(T), name = "t")
  ) + #lims(x=c(1,10), y = c(1,10))+
  theme_minimal() +
  theme(axis.title.y.left = element_blank()
        , axis.ticks.y.left=element_blank(), axis.text.y.left=element_blank()) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "t-s indices", subtitle="T=10") + coord_fixed()
  
p1 + p2 + plot_annotation(tag_levels = 'A', tag_prefix = "(", , tag_suffix = ")")
```

Note that the inner sum on $t$ gives $s-1$ pairs, so that the total number of pairs is

$$
\sum_{s=2}^T (s-1) = 1+2+3+\cdots (T-1) = \frac{T(T-1)}{2}
$$

or half of all $T^2$ index pairs $(t,s)$ less the diagonal $(s,s)$. From figure (B) we see that it sums the lower diagonal of all index pair values, less the diagonal.

This sum also counts the number of unique pairs $(s,t)$ where $2\le s\le T$ and $1\le t<s$ which is equivalent to counting the number of pairs where $1 \le t < s \le T$

### term 4:

The fourth term is $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t}b_{t+k}$ and again we will use the change of variables $s=t+k$.

With the change of variable, and with $s$ now the second index, the inner sum goes from $s=t+1$ to $s=T$ and the outer sum goes from $t=1$ to $t=T-1$. The $(t,s)$ indices used to sum $-\frac{1}{T^{2}}\sum_{t=1}^{T-1}\sum_{s=t+1}^{T}a_{t}b_{s}$ are shown in figure (B) below (for $T=10$).

In fact the $a_{t}b_{t+k}$ terms are the transpose of the terms $a_{t+k}b_{t}$ of figure (B) as can be seen in figure (C).

```{r}
#| label: second grid plot
#| echo: false
#| fig-align: center

p3 <- res |> 
  ggplot(aes(x=t,y=s))+
  geom_point(shape=0, color = "#21908C") +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(name = "t", breaks = 1:T) +
  scale_y_continuous(
    sec.axis = ggplot2::sec_axis(transform = ~., breaks = 2:(T), name = "s")
  ) + #lims(x=c(1,10), y = c(1,10))+
  theme_minimal() +
  theme(axis.title.y.left = element_blank(), axis.ticks.y.left=element_blank(), axis.text.y.left=element_blank()) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "t-s indices", subtitle="T=10") + coord_fixed()

p3 + plot_annotation(tag_levels = list('C'), tag_prefix = "(", , tag_suffix = ")")

```

So per terms 3 & 4, each cross-product ($a_i b_j,\,i\ne j$) appears exactly once with a negative sign, for a total of $T^2-T$ terms (since each of terms 3 & 4 has $\frac{T(T-1)}{2}$ terms.

So we are missing a set of $T$ diagonal terms from the total $T^2$ terms.

but the sum of the first two terms is $\frac{T-1}{T^2}\sum_{t=1}^{T}a_t b_t$, which provides the missing diagonal (i.e. $-\frac{1}{T^2}\sum_{t=1}^{T}a_t b_t$), so we have:

\begin{align}
\frac{T-1}{T^{2}}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T^{2}}\sum_{s=1}^{T}\sum_{t=1,\,t\ne s}^{T}a_{t}b_{s} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T^{2}}\sum_{s=1}^{T}\sum_{t=1}^{T}a_{t}b_{s} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T}\sum_{s=1}^{T}b_{s}\times\frac{1}{T}\sum_{t=1}^{T}a_{t} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\bar{b}\bar{a}
\end{align}

### compute example for covariance:

::: {.center style="data-latex"}
```{r}
#| label: compute test - covariance
#| echo: true
#| results: asis


set.seed(8740); T = 20
a <- 1 + rnorm(n = T)
b <- a + 4 + rnorm(n = T, sd = 2)

# calculate using difference equation
cov_diff <- 
  1:(T-1) |> purrr::map_vec(
    (\(k){ ((dplyr::lead(a,k) - a) * (dplyr::lead(b,k) - b)) |> sum(na.rm=TRUE) })
  ) |> sum(na.rm=TRUE) / (T*T)

# summarize results
tibble::tibble(x = a, y = b) |> 
  dplyr::mutate(x = x - mean(x), y = y - mean(y), prod = x*y) |> 
  dplyr::summarize(pop_cov = mean(prod) ) |> 
  tibble::add_column(cov_diff = cov_diff) |> 
  gt::gt() |> 
  gt::tab_header(
    title = "Covariance Calculations"
    , subtitle = stringr::str_glue("T={T}")) |> 
  gt::cols_width( everything() ~ gt::px(150) ) |> 
  gt::cols_label(
    pop_cov = gt::md("using product formula")
    , cov_diff = "using diff formula"
  ) |> 
  gt::tab_options(table.width = gt::pct(50), table.align = "center") |> 
  gtExtras::gt_theme_espn() #|> gt::as_latex() 
```
:::

\newpage

### compute example for covariance:

The difference formula for (population) covariance is

$\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}-a_{t}\right)\left(b_{t+k}-b_{t}\right)$

And the same expression holds for variance calculations: the population variance is calculated as the mean of the square of centered values, can also be calculated using all the differences between values.

```{r}
#| label: compute test - variance
#| echo: true
#| code-fold: true

a_var <- (a-mean(a))^2 |> mean()

# calculate using difference equation
var_diff <- 
  1:(T-1) |> purrr::map_vec(
    (\(k){ ((dplyr::lead(a,k) - a)^2) |> sum(na.rm=TRUE) })
  ) |> sum(na.rm=TRUE) / (T*T)

# summarize results
tibble::tibble("using product formula" = a_var, "using difference formula" = var_diff) |> 
  gt::gt() |> 
  gt::tab_header(
    title = "Variance Calculations"
    , subtitle = stringr::str_glue("with variable a | T={T}")
  ) |> 
  gt::tab_options(table.width = gt::pct(50), table.align = "center") |> 
  gtExtras::gt_theme_espn()
```
