---
title: "Covariance Decomposition"
monofont: 'Source Code Pro'
monofontoptions: 
  - Scale=0.55
---

```{r}
#| include: false
#| echo: false

# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(
viridis, ggplot2, patchwork, showtext, extrafont
)

sysfonts::font_add_google("Lato")
showtext::showtext_auto()
theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))
```

Here we show that the empirical covariance can be expressed as a difference equation.

For $\Delta_k a_t \equiv a_{t+k}-a_t$ (the difference) and $\bar{a}\equiv\frac{1}{T}\sum_{t=1}^T a_t$ (the mean)

Given sequences $\left\{a_t\right\}_{t=1}^T$ and $\left\{b_t\right\}_{t=1}^T$, we have

$$\frac{1}{T}\sum_{t=1}^T(a_t-\bar{a})(b_t-\bar{b})= \frac{1}{T^2}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k} \Delta_k a_t \Delta_k b_t$$

## Strategy

The empirical covariance can be written as follows:

$$
\begin{align*}
\frac{1}{T}\sum_{t=1}^{T}(a_{t}-\bar{a})(b_{t}-\bar{b}) & =\frac{1}{T}\sum_{t=1}^{T}\left(a_{t}b_{t}-\bar{a}b_{t}-a_{t}\bar{b}+\bar{a}\bar{b}\right)\\
 & =\left(\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}\right)-\bar{a}\bar{b}
\end{align*}
$$

so the proof strategy is to show that the difference equation can be reduced to the same form.

## Proof

Since $\Delta_k a_t \equiv a_{t+k}-a_t$ we can expand the difference equation as:

$$
\begin{align*}
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\Delta_{k}a_{t}\Delta_{k}b_{t} & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}-a_{t}\right)\left(b_{t+k}-b_{t}\right)\\
 & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}b_{t+k}+a_{t}b_{t}\right)\\
 & -\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t}b_{t+k}-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}
\end{align*}
$$ and we'll evaluate each of the 4 terms in turn.

### term 1:

Making a change of variable: $s=t+k$ and noting that when $t$ ranges from $1$ to $T−k$, $s$ ranges from $1+k$ to $T$, we have:

$$
\begin{align*}
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t+k} & =\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{s=k+1}^{T}a_{s}b_{s}\\
 & =\frac{1}{T^{2}}\sum_{s=2}^{T}\sum_{k=1}^{s-1}a_{s}b_{s}\\
 & =\frac{1}{T^{2}}\sum_{s=2}^{T}\left(s-1\right)a_{s}b_{s}
\end{align*}
$$

where in the second-last line we note that $a_s b_s$ doesn't depend on $k$ so we count each $a_s b_s$ exactly $s-1$ times for $s\in[2,T]$. Note that this is

$$
\frac{1}{T^{2}}\left(a_{2}b_{2}+2a_{3}b_{3}+3a_{4}b_{4}+\cdots++\left(T-2\right)a_{T-1}b_{T-1}+\left(T-1\right)a_{T}b_{T}\right)
$$

### term 2:

The second term is

$$
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_t b_t
$$ and it can be decomposed directly to:

$$
\frac{1}{T^{2}}\left((T-1)a_{1}b_{1}+(T-2)a_{2}b_{2}+(T-3)a_{3}b_{3}+\cdots+a_{T-1}b_{T-1}\right)
$$

So combining the first and second terms we get $\frac{T-1}{T^2}\sum_{t=1}^{T}a_t b_t$, and we're on our way towards showing that

$$
\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\Delta_{k}a_{t}\Delta_{k}b_{t} = \left(\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}\right)-\bar{a}\bar{b}
$$

### term 3:

The third term is $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}$ and again we will use the change of variables $s=t+k$.

Before changing variables, the outer sum goes from $k=1$ to $k=T-1$ and the inner sum goes from $t=1$ to $t=T-k$. The $(k,t)$ indices used in $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t+k}b_{t}$ are shown in figure (A) below (for $T=10$), along with the $s=t+k$ values.

With the change of variable, the outer sum goes from $s=t+k=2$ to $s=t+k=T$ and the inner sum goes from $t=1$ to $t=s-1$. The $(s,t)$ indices used to sum $-\frac{1}{T^{2}}\sum_{s=2}^{T}\sum_{t=1}^{s-1}a_{s}b_{t}$ are shown in figure (B) below (for $T=10$).

The first index ($s$) is along the horizontal axis, while the second index ($t$) is along the vertical axis.

```{r}
#| label: first grid plot
#| echo: true
#| code-fold: true
#| fig-align: center

T <- 10
res <- 1:(T-1) |> 
  purrr::map( 
    (\(k){ 1:(T-k) |> 
        purrr::map(\(t)data.frame("k"=k,"t"=t, "s"=t+k)) }
    ) 
  ) |> dplyr::bind_rows()

p1 <- res |> 
  ggplot(aes(x=k,y=t, label = as.character(s))) + geom_label(color = "#3B528B") +
  #geom_point(color = "#3B528B") +
  scale_x_continuous(breaks = 1:T) +
  scale_y_continuous(
    "t"
    , breaks = 1:T
    , labels = 1:T |> as.character()
  ) +
  theme_minimal()  +
  theme(axis.title.y.left = element_text(angle = 0, vjust = 0.5)) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "s values given k and t", subtitle="T=10")+ coord_fixed()

p2 <- res |> 
  ggplot(aes(x=s,y=t+1))+
  geom_point(shape=0, color = "#21908C") +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(name = "s", breaks = 1:T) +
  scale_y_continuous(
    sec.axis = ggplot2::sec_axis(transform = ~.-1, breaks = 1:(T), name = "t")
  ) + #lims(x=c(1,10), y = c(1,10))+
  theme_minimal() +
  theme(axis.title.y.left = element_blank()
        , axis.ticks.y.left=element_blank(), axis.text.y.left=element_blank()) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "t-s indices", subtitle="T=10") + coord_fixed()
  
p1 + p2 + plot_annotation(tag_levels = 'A', tag_prefix = "(", , tag_suffix = ")")

```

Note that the inner sum on $t$ gives $s-1$ pairs, so that the total number of pairs is

$$
\sum_{s=2}^T (s-1) = 1+2+3+\cdots (T-1) = \frac{T(T-1)}{2}
$$

or half of all $T^2$ index pairs $(t,s)$ less the diagonal $(s,s)$. From figure (B) we see that it sums the lower diagonal of all index pair values, less the diagonal.

This sum also counts the number of unique pairs $(s,t)$ where $2\le s\le T$ and $1\le t<s$ which is equivalent to counting the number of pairs where $1 \le t < s \le T$

### term 4:

The fourth term is $-\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}a_{t}b_{t+k}$ and again we will use the change of variables $s=t+k$.

With the change of variable, and with $s$ now the second index, the inner sum goes from $s=t+1$ to $s=T$ and the outer sum goes from $t=1$ to $t=T-1$. The $(t,s)$ indices used to sum $-\frac{1}{T^{2}}\sum_{t=1}^{T-1}\sum_{s=t+1}^{T}a_{t}b_{s}$ are shown in figure (B) below (for $T=10$).

In fact the $a_{t}b_{t+k}$ terms are the transpose of the terms $a_{t+k}b_{t}$ of figure (B) as can be seen in figure (C).

```{r}
#| label: second grid plot
#| echo: true
#| code-fold: true
#| fig-align: center

p3 <- res |> 
  ggplot(aes(x=t,y=s))+
  geom_point(shape=0, color = "#21908C") +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(name = "t", breaks = 1:T) +
  scale_y_continuous(
    sec.axis = ggplot2::sec_axis(transform = ~., breaks = 2:(T), name = "s")
  ) + #lims(x=c(1,10), y = c(1,10))+
  theme_minimal() +
  theme(axis.title.y.left = element_blank(), axis.ticks.y.left=element_blank(), axis.text.y.left=element_blank()) +
  theme(axis.title.y.right = element_text(angle = 0, vjust = 0.5), axis.ticks = element_blank()) +
  labs(title = "t-s indices", subtitle="T=10") + coord_fixed()

p3 + plot_annotation(tag_levels = list('C'), tag_prefix = "(", , tag_suffix = ")")

```

So per terms 3 & 4, each cross-product ($a_i b_j,\,i\ne j$) appears exactly once with a negative sign, for a total of $T^2-T$ terms (since each of terms 3 & 4 has $\frac{T(T-1)}{2}$ terms.

So we are missing a set of $T$ diagonal terms from the total $T^2$ terms.

but the sum of the first two terms is $\frac{T-1}{T^2}\sum_{t=1}^{T}a_t b_t$, which provides the missing diagonal, so we have:

$$
\begin{align*}
\frac{T-1}{T^{2}}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T^{2}}\sum_{s=1}^{T}\sum_{t=1,\,t\ne s}^{T}a_{t}b_{s} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T^{2}}\sum_{s=1}^{T}\sum_{t=1}^{T}a_{t}b_{s} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\frac{1}{T}\sum_{s=1}^{T}b_{s}\times\frac{1}{T}\sum_{t=1}^{T}a_{t} & =\\
\frac{1}{T}\sum_{t=1}^{T}a_{t}b_{t}-\bar{b}\bar{a}
\end{align*}
$$

### compute:

```{r}
#| label: compute test - covariance
#| echo: true
#| code-fold: true

set.seed(8740); T = 20
a <- 1 + rnorm(n = T)
b <- a + 4 + rnorm(n = T, sd = 2)

# calculate using difference equation
cov_diff <- 
  1:(T-1) |> purrr::map_vec(
    (\(k){ ((dplyr::lead(a,k) - a) * (dplyr::lead(b,k) - b)) |> sum(na.rm=TRUE) })
  ) |> sum(na.rm=TRUE) / (T*T)

# summarize results
tibble::tibble(x = a, y = b) |> 
  dplyr::mutate(x = x - mean(x), y = y - mean(y), prod = x*y) |> 
  dplyr::summarize(pop_cov = mean(prod), smpl_cov = sum(prod)/(dplyr::n()-1) ) |> 
  tibble::add_column(
    calc_cov = cov(a,b) # built in covariance 
    , cov_diff = cov_diff
  ) |> 
  gt::gt() |> 
  gt::tab_header(title = "Covariances", subtitle = stringr::str_glue("T={T}")) |> 
  gtExtras::gt_theme_espn()
```

$\frac{1}{T^{2}}\sum_{k=1}^{T-1}\sum_{t=1}^{T-k}\left(a_{t+k}-a_{t}\right)\left(b_{t+k}-b_{t}\right)$

And the same expression hold for variance calculations: the population variance, calculated as the mean of the square of centered values, can also be calculated using all the differences between values.

```{r}
#| label: compute test - variance
#| echo: true
#| code-fold: true

a_var <- (a-mean(a))^2 |> mean()

# calculate using difference equation
var_diff <- 
  1:(T-1) |> purrr::map_vec(
    (\(k){ ((dplyr::lead(a,k) - a)^2) |> sum(na.rm=TRUE) })
  ) |> sum(na.rm=TRUE) / (T*T)

# summarize results
tibble::tibble("population variance" = a_var, "by difference calculation" = var_diff) |> 
  gt::gt() |> 
  gt::tab_header(title = "Variances", subtitle = stringr::str_glue("variable a | T={T}")) |> 
  gtExtras::gt_theme_espn()
```

## Implication for regression coefficients

From FWL we know we can write our regression coefficient of interest as the ratio of a covariance over a variance, and so we can also express our coefficient in terms of differences.

$$
\begin{align*}
Y_{it} & =\alpha_{i}+\lambda_{t}+\delta\cdot D_{it};\;i\in\left\{ 0,1\right\} ,\,t\in\left\{ 0,1\right\} \\
\Delta_{1}Y_{it} & =\Delta_{1}\lambda_{t}+\delta\cdot\Delta_{1}D_{it}\\
Y_{i1}^{0}-Y_{i0}^{0} & =\lambda\left(1-0\right)+\delta\left(0-0\right)\\
Y_{i1}^{1}-Y_{i0}^{1} & =\lambda\left(1-0\right)+\delta\left(1-0\right)
\end{align*}
$$

```{r}
#| label: TWFE example
#| echo: true
#| code-fold: true
#| message: false
#| warning: false

# TWFE Example with 40 samples
# Setting seed for reproducibility
set.seed(123)

# Parameters
n_units <- 280        # Number of units/entities
n_periods <- 2      # Number of time periods
total_samples <- n_units * n_periods  # Total observations (40)

# Unit and time fixed effects
unit_effects <- rnorm(n_units, mean = 5, sd = 2)  # Unit-specific effects
time_effects <- rnorm(n_periods, mean = 0, sd = 1)  # Time-specific effects
epsilon      <- rnorm(total_samples, mean = 0, sd = 1)  # Error term

# True treatment effect
true_effect <- 2.5

df <- tibble::tibble(
  unit = rep(1:n_units, each = n_periods)
  , time = rep(1:n_periods, times = n_units)
  , treatment =
    dplyr::case_when((unit >= n_units/2) & (time == 2) ~ 1, TRUE ~ 0)
  , y = unit_effects[unit] +                  # Unit fixed effects
    time_effects[time] +                      # Time fixed effects
    true_effect * treatment +                 # Treatment effect
    epsilon 
  , unit_effect = unit_effects[unit]
  , time_effect = time_effects[time]
)

# panel data estimators
twfe_model <- plm::plm(y ~ treatment, 
                 data = df |> as.data.frame(), 
                 index = c("unit", "time"), 
                 model = "within", 
                 effect = "twoways")

# linear model
twfe_lm_model <- lm(y ~ 1 + time + treatment, data = df)
# lm(y ~ 1 + factor(unit) + factor(time) + factor(treatment), data = df)

# plm::plm(y ~ + treatment, data = df, index = c("unit", "time"), model = "random", effect = "twoways")

# Display model results
g <- twfe_model |> broom::tidy() |> dplyr::mutate(model = 'twfe_plm_model', .before = 1) |> 
  dplyr::bind_rows(
    twfe_lm_model |> broom::tidy() |> dplyr::mutate(model = 'twfe_lm_model', .before = 1)
  ) |> dplyr::group_by(model) |> 
  gt::gt("term") |> 
  gt::fmt_number(columns = -c(model,term), decimals = 3) |> 
  gt::tab_header(title = "TWFE Estimates", subtitle = "using lm and plm") |> 
  gtExtras::gt_theme_espn()

df_plot_data <- df |> dplyr::group_by(unit) |> 
  dplyr::group_map(
    .f = 
      ~dplyr::mutate(
        .x, treated = dplyr::case_when(1 %in% treatment ~ "treated", TRUE ~ "control" )
      )
    , .keep = TRUE
  ) |> 
  dplyr::bind_rows() |> dplyr::group_by(time, treated) |> 
  dplyr::summarize(mean_y = mean(y))

# p1 <- df_plot_data |> 
#   ggplot(aes(x = time, y = mean_y, group = treated, color = treated)) +
#   geom_line(linewidth = 1) +
#   geom_point(size = 3) +
#   labs(title = "Difference-in-Differences Visual",
#        subtitle = paste("True treatment effect =", true_effect),
#        x = "Time Period", y = "Mean Outcome", color = "Group") +
#   scale_x_continuous(breaks = c(1, 2)) +
#   theme_minimal() +
#   theme(legend.position = "none")
  

# Plot group means over time
# p1 <- ggplot(plot_data, aes(x = time, y = mean_y, group = group, color = group)) +
#   geom_line(linewidth = 1) +
#   geom_point(size = 3) +
#   labs(title = "Difference-in-Differences Visual",
#        subtitle = paste("True treatment effect =", true_effect),
#        x = "Time Period", y = "Mean Outcome", color = "Group") +
#   scale_x_continuous(breaks = c(1, 2)) +
#   theme_minimal()

# # Plot individual unit trajectories
# p2 <- df |> dplyr::group_by(unit) |> 
#   dplyr::group_map(
#     .f = ~dplyr::mutate(.x, treated = dplyr::case_when(1 %in% treatment ~ "treated", TRUE ~ "control" ))
#     , .keep = TRUE
#   ) |> 
#   dplyr::bind_rows() |> 
#   rsample::group_initial_split(group=unit, prop = 0.2, strata = treated) |> 
#   rsample::training() |> 
#   ggplot(aes(x = time, y = y, group = unit, 
#                color = treated)) + # factor(ifelse(unit <= T/2, "Control", "Treated"))
#   geom_line(alpha = 0.3,  ) +
#   geom_point() +
#   # plot_data, aes(x = time, y = mean_y, group = group, color = group)) +
#   labs(title = "Individual Unit Trajectories",
#        x = "Time Period", y = "Outcome", color = "Group") +
#   scale_x_continuous(breaks = c(1, 2)) +
#   #theme_set(theme_bw(base_size = 12) + theme(legend.position = "top"))
#   theme_minimal()

p3 <- df |> dplyr::group_by(unit) |> 
  dplyr::group_map(
    .f = ~dplyr::mutate(.x, treated = dplyr::case_when(1 %in% treatment ~ "treated", TRUE ~ "control" ))
    , .keep = TRUE
  ) |> 
  dplyr::bind_rows() |> 
  rsample::group_initial_split(group=unit, prop = 0.2, strata = treated) |> 
  rsample::training() |> 
  ggplot(aes(x = time, y = y, group = unit, color = treated)) + 
  geom_line(alpha = 0.8, linetype=3 ) +
  geom_point(alpha = 0.8, shape = 15) +
  geom_line(
    data=df_plot_data, aes(x = time, y = mean_y, group = treated, color = treated), linewidth = 1
  ) +
  labs(x = "Time Period", y = "Outcome", color = "Group") +
  scale_x_continuous(breaks = c(1, 2)) +
  theme_minimal() +
  theme(legend.title = element_blank()) 

p3 + wrap_table(g, panel="full", space = "free_y") + plot_annotation(
  title = 'Individual Unit Trajectories (dotted), Mean Values (solid), and Model Estimates',
  subtitle = 
    stringr::str_glue("True treatment effect = {true_effect} | {n_units} units | {n_periods} periods")
)
```

```{r}
#| label: 4 means and 3 subtractions
#| echo: true
#| code-fold: true 

# Calculate means by group and time
group_means <- stats::aggregate(y ~ treatment + time, data = df, FUN = mean) |> 
  dplyr::mutate(group = dplyr::case_when(treatment == 1 ~ "Treated", TRUE ~ "Control"))

group_means |> tidyr::pivot_wider(names_from = group, values_from = y) |> 
  tibble::add_row(treatment = 1, time = 1, .before = 3) |> 
  dplyr::mutate(
    Treated = dplyr::case_when(dplyr::row_number() == 3 ~ mean(df$y[df$unit %in% 11:20 & df$time == 1]), TRUE ~ Treated)
    , Treated = dplyr::lead(Treated,2)) |> 
  dplyr::reframe(dplyr::across(.cols=Control:Treated, .fns = ~dplyr::lead(.x)-.x)) |> 
  tidyr::drop_na() |> 
  dplyr::mutate(diff = Treated - Control) |> 
  gt::gt() |> 
  gt::fmt_number(decimals = 3) |> 
  gt::tab_header(title = "Manually calculated difference-in-differences") |> 
  gt::cols_label(
    Control = "control mean diffs",
    Treated = "treated mean diffs",
    diff = "Treated - Control"
  ) |> 
  gtExtras::gt_theme_espn()

```

```{r}
#| label: using first diffs
#| echo: false
#| eval: true
#| code-fold: true


df_delta <- df |> dplyr::group_by(unit) |> 
  dplyr::mutate(across(time:y, .fns = list(diff = ~(dplyr::lead(.x)-.x) ))) |> 
  tidyr::drop_na() |> 
  dplyr::ungroup() |> 
  dplyr::select(ends_with("diff")) |> 
  lm(y_diff ~ time_diff + treatment_diff, data = _)

df_delta |> broom::tidy() |> tidyr::drop_na() |> 
  gt::gt() |> 
  gt::fmt_number(decimals = 3) |> 
  gt::tab_header(title = "Estimated by differences") |> 
  gtExtras::gt_theme_espn()

# df |> panelView::panelview(Y="y", D="treatment", index = c("unit","time"))|>class()


```

```{r}
#| label: XX
#| echo: false
#| eval: false
#| code-fold: true

# TWFE Example with 40 samples and 2 time periods
# Setting seed for reproducibility
set.seed(123)

# Parameters
n_units <- 20       # Number of units/entities (increased from 8)
n_periods <- 2      # Number of time periods (reduced from 5)
total_samples <- n_units * n_periods  # Total observations (40)

# Create balanced panel data
create_twfe_data <- function() {
  # Create unit and time indices
  unit_id <- rep(1:n_units, each = n_periods)
  time_id <- rep(1:n_periods, times = n_units)
  
  # Unit and time fixed effects
  unit_effects <- rnorm(n_units, mean = 5, sd = 2)  # Unit-specific effects
  time_effects <- rnorm(n_periods, mean = 0, sd = 1)  # Time-specific effects
  
  # Assign treatment (difference-in-differences design)
  # Half of the units are treated in the second period
  treated_units <- 11:20  # Units 11-20 get treated in period 2
  treatment <- sapply(1:length(unit_id), function(i) {
    unit_id[i] %in% treated_units && time_id[i] == 2
  }) * 1
  
  # True treatment effect
  true_effect <- 2.5
  
  # Generate outcome with unit fixed effects, time fixed effects, and treatment effect
  epsilon <- rnorm(total_samples, mean = 0, sd = 1)  # Error term
  y <- unit_effects[unit_id] +                       # Unit fixed effects
       time_effects[time_id] +                       # Time fixed effects
       true_effect * treatment +                     # Treatment effect
       epsilon                                       # Error term
  
  # Create dataframe
  df <- data.frame(
    unit = unit_id,
    time = time_id,
    treatment = treatment,
    y = y,
    unit_effect = unit_effects[unit_id],
    time_effect = time_effects[time_id]
  )
  
  return(list(df = df, true_effect = true_effect))
}

# Generate the data and store
twfe_data <- create_twfe_data()
df <- twfe_data$df
true_effect <- twfe_data$true_effect

# Display the data
print("First few rows of the dataset:")
head(df, 10)
print(paste("Total observations:", nrow(df)))
print(paste("True treatment effect:", true_effect))

# Display treatment patterns
print("Treatment pattern by unit and time:")
treatment_pattern <- reshape(df[, c("unit", "time", "treatment")], 
                            idvar = "unit", 
                            timevar = "time", 
                            direction = "wide")
colnames(treatment_pattern) <- c("unit", "t1", "t2")
print(treatment_pattern)

# Calculate means by group and time
group_means <- aggregate(y ~ treatment + time, data = df, FUN = mean)
group_means$group <- ifelse(group_means$treatment == 1, "Treated", "Control")
print("Group means by treatment status and time:")
print(group_means)

# Manually calculate difference-in-differences
control_t1 <- group_means$y[group_means$group == "Control" & group_means$time == 1]
control_t2 <- group_means$y[group_means$group == "Control" & group_means$time == 2]
treated_t1 <- group_means$y[group_means$group == "Treated" & group_means$time == 1]
treated_t2 <- group_means$y[group_means$group == "Treated" & group_means$time == 2]

# Note: For t1, treated_t1 will actually be from the "Control" group since treatment=0 for all units at t=1
treated_t1 <- mean(df$y[df$unit %in% 11:20 & df$time == 1])

did_estimate <- (treated_t2 - treated_t1) - (control_t2 - control_t1)
print(paste("Manual DiD estimate:", round(did_estimate, 4)))

# Estimate the TWFE model
library(plm)
twfe_model <- plm::plm(y ~ treatment, 
                 data = df, 
                 index = c("unit", "time"), 
                 model = "within", 
                 effect = "twoways") |> broom::tidy()

# Display model results
print("TWFE Model Results:")
summary(twfe_model)

# Create a more detailed 2x2 DiD table
did_table <- matrix(c(control_t1, control_t2, treated_t1, treated_t2), ncol=2, byrow=TRUE)
rownames(did_table) <- c("Control", "Treated")
colnames(did_table) <- c("t=1", "t=2")
print("Difference-in-Differences Table:")
print(did_table)
print(paste("First difference (Treated): ", round(treated_t2 - treated_t1, 4)))
print(paste("First difference (Control): ", round(control_t2 - control_t1, 4)))
print(paste("Difference-in-differences: ", round(did_estimate, 4)))

# Visualize the data
library(ggplot2)

# Prepare data for plotting group means
plot_data <- data.frame(
  time = c(1, 2, 1, 2),
  group = c("Control", "Control", "Treated", "Treated"),
  mean_y = c(control_t1, control_t2, treated_t1, treated_t2)
)

# Plot group means over time
p1 <- ggplot(plot_data, aes(x = time, y = mean_y, group = group, color = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(title = "Difference-in-Differences Visual",
       subtitle = paste("True treatment effect =", true_effect),
       x = "Time Period", y = "Mean Outcome", color = "Group") +
  scale_x_continuous(breaks = c(1, 2)) +
  theme_minimal()

# Plot individual unit trajectories
p2 <- df |> dplyr::group_by(unit) |> 
  dplyr::group_map(
    .f = ~dplyr::mutate(.x, treated = dplyr::case_when(1 %in% treatment ~ "treated", TRUE ~ "control" ))
    , .keep = TRUE
  ) |> 
  dplyr::bind_rows() |> 
  rsample::group_initial_split(group=unit, prop = 0.2, strata = treated) |> 
  rsample::training() |> 
  ggplot(aes(x = time, y = y, group = unit, 
               color = treated)) + # factor(ifelse(unit <= T/2, "Control", "Treated"))
  geom_line(alpha = 0.3,  ) +
  geom_point() +
  labs(title = "Individual Unit Trajectories",
       x = "Time Period", y = "Outcome", color = "Group") +
  scale_x_continuous(breaks = c(1, 2)) +
  theme_minimal()

p1+p2+ plot_layout(axes = "collect_y")
```

```{r}
# TWFE Example with 40 samples, 3 time periods, treatment in period 2 only
# Setting seed for reproducibility
# TWFE Example with 100 units, 3 time periods, treatment in period 2 only
# Setting seed for reproducibility
set.seed(123)

# Parameters
n_units <- 100       # Number of units/entities
n_periods <- 3       # Number of time periods
total_samples <- n_units * n_periods  # Total observations (300)

# Create balanced panel data
create_twfe_data <- function() {
  # Create unit and time indices
  unit_id <- rep(1:n_units, each = n_periods)
  time_id <- rep(1:n_periods, times = n_units)
  
  # Unit and time fixed effects
  unit_effects <- rnorm(n_units, mean = 5, sd = 2)  # Unit-specific effects
  time_effects <- c(0, 0.5, 1)  # Time trend
  
  # Assign treatment (only in period 2)
  # Half of the units are treated, but only in period 2
  treated_units <- 51:100  # Units 51-100 get treated in period 2 only
  treatment <- sapply(1:length(unit_id), function(i) {
    unit_id[i] %in% treated_units && time_id[i] == 2
  }) * 1
  
  # True treatment effect
  true_effect <- 2.5
  
  # Generate outcome with unit fixed effects, time fixed effects, and treatment effect
  epsilon <- rnorm(total_samples, mean = 0, sd = 1)  # Error term
  y <- unit_effects[unit_id] +                       # Unit fixed effects
       time_effects[time_id] +                       # Time fixed effects
       true_effect * treatment +                     # Treatment effect
       epsilon                                       # Error term
  
  # Create dataframe
  df <- data.frame(
    unit = unit_id,
    time = time_id,
    treatment = treatment,
    y = y,
    unit_effect = unit_effects[unit_id],
    time_effect = time_effects[time_id]
  )
  
  # Add treatment group indicator
  df$group <- ifelse(df$unit %in% treated_units, "Treated", "Control")
  
  return(list(df = df, true_effect = true_effect))
}

# Generate the data and store
twfe_data <- create_twfe_data()
df <- twfe_data$df
true_effect <- twfe_data$true_effect

# Basic data summary
print(paste("Total observations:", nrow(df)))
print(paste("Number of units:", length(unique(df$unit))))
print(paste("Number of time periods:", length(unique(df$time))))
print(paste("True treatment effect:", true_effect))

# Treatment summary
treatment_summary <- table(df$group, df$time, df$treatment)
print("Treatment pattern:")
print(treatment_summary)

# Calculate group means by treatment status and time
group_means <- aggregate(y ~ group + time, data = df, FUN = mean)
print("Group means by group and time period:")
print(group_means)

# Calculate treatment effect using a regression
library(plm)
twfe_model <- plm(y ~ treatment, 
                 data = df, 
                 index = c("unit", "time"), 
                 model = "within", 
                 effect = "twoways")

# Display model results
print("TWFE Model Results:")
summary_result <- summary(twfe_model)
print(summary_result)

# Calculate manual DiD estimate for period 2
get_did_estimate <- function(df) {
  # Pre-period difference (time=1)
  pre_diff <- mean(df$y[df$group == "Treated" & df$time == 1]) - 
              mean(df$y[df$group == "Control" & df$time == 1])
  
  # Treatment period difference (time=2)
  treat_diff <- mean(df$y[df$group == "Treated" & df$time == 2]) - 
                mean(df$y[df$group == "Control" & df$time == 2])
  
  # DiD estimate
  did_est <- treat_diff - pre_diff
  
  return(did_est)
}

manual_did <- get_did_estimate(df)
print(paste("Manual DiD estimate:", round(manual_did, 4)))

# Calculate event study estimates
event_study_model <- lm(y ~ factor(unit) + factor(time) + 
                           group:factor(time), data = df)

# Extract coefficients for group:time interactions
es_coefs <- coef(event_study_model)[grepl("group.*:factor\\(time\\)", names(coef(event_study_model)))]
print("Event study coefficients (relative to period 1):")
print(es_coefs)

# Visualize the data
library(ggplot2)

# Plot group means over time
ggplot(group_means, aes(x = time, y = y, group = group, color = group)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "lightblue") +
  annotate("text", x = 2, y = max(group_means$y), 
           label = "Treatment Period", vjust = -1) +
  labs(title = "Effect of Treatment in Period 2 Only",
       subtitle = paste("True treatment effect =", true_effect),
       x = "Time Period", y = "Mean Outcome", color = "Group") +
  scale_x_continuous(breaks = 1:3) +
  theme_minimal()

# Calculate and plot the distribution of treatment effects
df_period2 <- df[df$time == 2,]
df_period1 <- df[df$time == 1,]

# Merge period 1 data to period 2
df_period2$y_pre <- df_period1$y[match(df_period2$unit, df_period1$unit)]
df_period2$y_diff <- df_period2$y - df_period2$y_pre

# Plot distribution of changes by group
ggplot(df_period2, aes(x = y_diff, fill = group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = mean(df_period2$y_diff[df_period2$group == "Treated"]), 
             color = "red", linetype = "dashed") +
  geom_vline(xintercept = mean(df_period2$y_diff[df_period2$group == "Control"]), 
             color = "blue", linetype = "dashed") +
  labs(title = "Distribution of Changes from Period 1 to 2",
       subtitle = paste("Mean difference (Treated - Control):", 
                       round(mean(df_period2$y_diff[df_period2$group == "Treated"]) - 
                             mean(df_period2$y_diff[df_period2$group == "Control"]), 4)),
       x = "Change in Outcome", y = "Density", fill = "Group") +
  theme_minimal()

# Plot a subset of individual trajectories (20 units for clarity)
sampled_units <- c(sample(1:50, 10), sample(51:100, 10))
df_subset <- df[df$unit %in% sampled_units,]

ggplot(df_subset, aes(x = time, y = y, group = unit, 
                     color = group)) +
  geom_line(alpha = 0.7) +
  geom_point(aes(shape = factor(treatment)), size = 3) +
  scale_shape_manual(values = c(16, 8), name = "Treated") +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymin = -Inf, ymax = Inf, 
           alpha = 0.1, fill = "lightblue") +
  labs(title = "Sample of Individual Unit Trajectories",
       subtitle = "Treatment occurs only in period 2 for treated units",
       x = "Time Period", y = "Outcome", color = "Group") +
  scale_x_continuous(breaks = 1:3) +
  theme_minimal()

# Perform placebo test comparing period 1 to period 3 changes (no treatment in either)
df_period3 <- df[df$time == 3,]
df_period1$y_post <- df_period3$y[match(df_period1$unit, df_period3$unit)]
df_period1$y_diff <- df_period1$y_post - df_period1$y

placebo_diff <- mean(df_period1$y_diff[df_period1$group == "Treated"]) - 
                mean(df_period1$y_diff[df_period1$group == "Control"])

print(paste("Placebo test (period 1 to 3 difference):", round(placebo_diff, 4)))

# Test for heterogeneous treatment effects
df_treated_p2 <- df[df$treatment == 1,]
df_treated_p2$unit_effect_quantile <- cut(df_treated_p2$unit_effect, 
                                         breaks = quantile(df_treated_p2$unit_effect, probs = seq(0, 1, 0.25)),
                                         labels = c("Q1", "Q2", "Q3", "Q4"), 
                                         include.lowest = TRUE)

# Calculate treatment effect by unit effect quantile
df_treated_p1 <- df[df$group == "Treated" & df$time == 1,]
df_treated_p1$unit_effect_quantile <- cut(df_treated_p1$unit_effect, 
                                         breaks = quantile(df_treated_p1$unit_effect, probs = seq(0, 1, 0.25)),
                                         labels = c("Q1", "Q2", "Q3", "Q4"), 
                                         include.lowest = TRUE)

df_treated_p2$y_pre <- df_treated_p1$y[match(df_treated_p2$unit, df_treated_p1$unit)]
df_treated_p2$effect_estimate <- df_treated_p2$y - df_treated_p2$y_pre

# Get control trend
control_trend <- mean(df$y[df$group == "Control" & df$time == 2]) - 
                mean(df$y[df$group == "Control" & df$time == 1])

df_treated_p2$effect_adjusted <- df_treated_p2$effect_estimate - control_trend

het_effects <- aggregate(effect_adjusted ~ unit_effect_quantile, 
                        data = df_treated_p2, FUN = mean)

print("Heterogeneous treatment effects by unit effect quantile:")
print(het_effects)

# Plot heterogeneous effects
ggplot(het_effects, aes(x = unit_effect_quantile, y = effect_adjusted)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = true_effect, linetype = "dashed", color = "red") +
  labs(title = "Heterogeneous Treatment Effects",
       subtitle = "By Unit Effect Quantile",
       x = "Unit Effect Quantile", y = "Estimated Treatment Effect") +
  theme_minimal()

# Check power of the test
power_analysis <- function(df, n_simulations = 1000) {
  # Extract model parameters
  sigma <- sd(residuals(twfe_model))
  effect_sizes <- seq(0, 5, 0.5)
  
  # Store power results
  power_results <- data.frame(
    effect_size = effect_sizes,
    power = NA
  )
  
  for (i in 1:length(effect_sizes)) {
    # Track significant results
    significant_results <- 0
    
    for (j in 1:n_simulations) {
      # Create simulated data with this effect size
      df_sim <- df
      df_sim$y <- df_sim$unit_effect + df_sim$time_effect + 
                 effect_sizes[i] * df_sim$treatment + 
                 rnorm(nrow(df_sim), mean = 0, sd = sigma)
      
      # Run the model
      model_sim <- plm(y ~ treatment, 
                      data = df_sim, 
                      index = c("unit", "time"), 
                      model = "within", 
                      effect = "twoways")
      
      # Check if result is significant at 5% level
      p_value <- summary(model_sim)$coefficients[1, "Pr(>|t|)"]
      if (p_value < 0.05) {
        significant_results <- significant_results + 1
      }
    }
    
    # Calculate power
    power_results$power[i] <- significant_results / n_simulations
  }
  
  return(power_results)
}

# Run a smaller number of simulations for demonstration
power_results <- power_analysis(df, n_simulations = 100)
print("Power analysis results:")
print(power_results)

# Plot power curve
ggplot(power_results, aes(x = effect_size, y = power)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  geom_vline(xintercept = true_effect, linetype = "dashed", color = "blue") +
  labs(title = "Power Analysis",
       subtitle = "Power to detect different effect sizes",
       x = "Effect Size", y = "Power") +
  theme_minimal()
```

```{r}
#| label: using diffs - three periods
#| echo: false
#| eval: true
#| code-fold: true


df_delta <- df |> dplyr::group_by(unit) |> 
  dplyr::mutate(
    across(
      time:y
      , .fns = list(diff = ~(dplyr::lead(.x)-.x) )
      , .names = "{.col}_{.fn}"
    )
  ) |> 
  tidyr::drop_na() |> 
  dplyr::ungroup() |> 
  dplyr::select(ends_with("diff")) 

df_all_delta <- purrr::reduce(
  .x = 1:2
  , .f = (\(x,k){
    x |>
      dplyr::mutate(
        across(
          time:y
          , .fns = list(diff = ~(dplyr::lead(.x, n=k)-.x) )
          , .names = "{.col}_{.fn}_{k}"
        )
      )
  })
  , .init = df |> dplyr::group_by(unit)
)

1:2 |> 
  purrr::map(
    .f = 
      (\(k){
        fml <- formula(stringr::str_glue("y_diff_{k} ~ time_diff_{k} + treatment_diff_{k}"))
        df_all_delta |> 
          dplyr::select( ends_with(stringr::str_glue("diff_{k}")) ) |> 
          dplyr::ungroup() |>
          tidyr::drop_na() |> 
          lm(fml, data = _) |> 
          broom::tidy()
      })
  ) |> 
  dplyr::bind_rows()

|> 
  lm(y_diff ~ time_diff + treatment_diff, data = _)






df_delta |> broom::tidy() |> tidyr::drop_na() |> 
  gt::gt() |> 
  gt::fmt_number(decimals = 3) |> 
  gt::tab_header(title = "Estimated by differences") |> 
  gtExtras::gt_theme_espn()

```

