---
title: "Decompositions"
format: pdf
monofontoptions: 
  - Scale=0.55
---

```{r}
#| include: false
#| echo: false

# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(
  magrittr, tidyverse, malcolmbarrett/causalworkshop, ggplot2, patchwork, AER
)

theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))
```

# Mind the Gap

When faced with a gap in mean outcomes between two groups, researchers frequently examine how much of the gap can be explained by differences in observable characteristics.

The simple approach is to estimate the pooled regression including an indicator variable for group membership as well as the other observable characteristics, interpreting the coefficient on the group indicator as the unexplained component.

The Oaxaca-Blinder (O-B) decomposition represents an alternative approach

## Oaxaca-Blinder decomposition

Consider a categorical (or dummy) variable $d$ that splits our dataset into two groups.

In this case we can run regressions of the form $y=X\beta+\epsilon$ to estimate the the mean difference between groups, as follows

\begin{align*}
\mathbb{E}\left[y^{0}\right] & =\mathbb{E}\left[X^{(0)}\right]\beta_{0};\;\text{group }d=0\\
\mathbb{E}\left[y^{1}\right] & =\mathbb{E}\left[X^{(1)}\right]\beta_{1};\;\text{group }d=1
\end{align*}

Alternatively

\begin{align*}
\bar{y}_0 & =\bar{X}_0\beta_{0};\;\text{group }d=0\\
\bar{y}_1 & =\bar{X}_1\beta_{1};\;\text{group }d=1
\end{align*}

Then the mean difference in outcomes is:

\begin{align*}
\mathbb{E}\left[y^{1}\right]-\mathbb{E}\left[y^{0}\right] & =\mathbb{E}\left[X^{(1)}\right]\beta_{1}-\mathbb{E}\left[X^{(0)}\right]\beta_{0}\\
 & =\left(\mathbb{E}\left[X^{(1)}\right]-\mathbb{E}\left[X^{(0)}\right]\right)\beta_{1}+\mathbb{E}\left[X^{(0)}\right]\left(\beta_{1}-\beta_{0}\right)\\
 & =\left(\mathbb{E}\left[X^{(1)}\right]-\mathbb{E}\left[X^{(0)}\right]\right)\beta_{0}-\mathbb{E}\left[X^{(1)}\right]\left(\beta_{0}-\beta_{1}\right)
\end{align*}

Alternatively

\begin{align*}\label{eq:gap}\tag{1}
\bar{y}_1-\bar{y}_0 & =\bar{X}_1\beta_{1}-\bar{X}_0\beta_{0}\\
 & =\left(\bar{X}_1-\bar{X}_0\right)\beta_{1}+\bar{X}_0\left(\beta_{1}-\beta_{0}\right)\\
 & =\left(\bar{X}_1-\bar{X}_0\right)\beta_{0}-\bar{X}_1\left(\beta_{0}-\beta_{1}\right)
\end{align*} 

Where: $\left( \bar{X}_1-\bar{X}_0\right)\beta_{1}$ is the "explained" component (differences in characteristics), and $\bar{X}_0\left(\beta_{1}-\beta_{0}\right)$) is the "unexplained" component (differences in returns to characteristics), where $\bar{X}_0$ is the baseline (this decomposition is not unique, as you can see from the third line of equation (\ref{eq:gap})

and we define

\begin{align*}
\text{Gap}^{1} & =\bar{X}_0\left(\beta_{1}-\beta_{0}\right)\\
\text{Gap}^{0} & =\bar{X}_1\left(\beta_{0}-\beta_{1}\right)\\
\text{Gap}^{\text{OLS}} & =\delta_{d};\;\text{where }y=\delta_{0}+\delta_{d}d+\delta_{1}X+\epsilon
\end{align*}

where the strong assumptions that the model is properly specified and that coefficients are equal across groups, a sensible definition of the population unexplained gap is $\delta_d$.

Example

data:

::: {style="font-size: small"}

\AddToHookNext{env/Highlighting/begin}{\tiny}
```{r}
#| eval: false
#| code-overflow: wrap
# https://cran.r-project.org/web/packages/oaxaca/vignettes/oaxaca.pdf
# https://www.worldbank.org/content/dam/Worldbank/document/HDN/Health/HealthEquityCh12.pdf
# https://pmc.ncbi.nlm.nih.gov/articles/PMC8343972/
# https://www.sciencedirect.com/science/article/abs/pii/S0169721811004072
# https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2528391
# https://link.springer.com/article/10.1186/s12982-021-00100-9
# https://giacomovagni.com/blog/2023/oaxaca/
# https://journals.sagepub.com/doi/pdf/10.1177/1536867X0800800401
# https://ocw.mit.edu/courses/14-662-labor-economics-ii-spring-2015/resources/mit14_662s15_lecnotes1/

# Load HMDA data
data('PSID1982', package = "AER")
hmda <- PSID1982

# data("nswcps", package = "hettreatreg")

# Prepare data for analysis
# Looking at income differences between racial groups
lending_data <- hmda |> 
  dplyr::mutate(
    minority = factor(race != "white"),
    log_income = log(income)
  ) |> 
  dplyr::select(log_income, minority, education, hrat, ccred, mcred, pubrec)
```
\AddToHookNext{env/Highlighting/begin}{\tiny}
```{r}
#| eval: false
#| code-overflow: wrap
# Fit separate regressions
model_a <- 
  lm(wage ~ education + experience, data = dat |> dplyr::filter(union=='yes'))
model_b <- 
  lm(wage ~ education + experience, data = dat |> dplyr::filter(union=='no'))

# Get mean characteristics (including intercept)
X_mean_a <- 
  c(1, 
    colMeans( 
    dat |> 
      dplyr::filter(union=='yes') |> 
      dplyr::select(education, experience)
    ) 
  )
X_mean_b <- 
  c(1, 
    colMeans( 
      dat |> 
        dplyr::filter(union=='no') |> 
        dplyr::select(education, experience) 
    ) 
  )

# Get coefficients
beta_a <- coef(model_a)
beta_b <- coef(model_b)

# Calculate decomposition
tibble::tibble(
  explained = sum((X_mean_a - X_mean_b) * beta_a)
  , unexplained = sum(X_mean_b * (beta_a - beta_b))
  , total_gap <- explained + unexplained
)

# Perform Oaxaca-Blinder decomposition
decomp <- 
  oaxaca::oaxaca(
    wage ~ education + experience | union
    , data = 
      dat |> 
      dplyr::mutate(union = dplyr::case_when(union=='yes'~0, TRUE ~1))
  )

```
:::

We an define another measure $\text{Gap}^{p}$ as

\begin{align*}
\bar{y}_1-\bar{y}_0 & =\left(\bar{X}_1-\bar{X}_0\right)\hat{\beta}^{p}+\text{Gap}^p\\
\text{Gap}^p & =\bar{X}_1\left(\hat{\beta}_1-\hat{\beta}^{p}\right)+\bar{X}_0\left(\hat{\beta}^{p}-\hat{\beta}^{0}\right)
\end{align*}

where $\hat{\beta}^{p}$ is the coefficient from the pooled regression of $y$ on $X$.

An O-B unexplained gap can always be written as the difference in overall mean outcomes minus the difference in predicted mean outcomes, and both of these differences can be denoted by linear projections.

A general expression for an O-B unexplained gap is

\begin{align*}
\text{Gap} & =\left[\bar{y}_{1}-\bar{y}_{0}\right]-\left[\hat{\beta}\left(\bar{x}_{1}-\bar{x}_{0}\right)\right]\\
 & =\text{b}(y\vert d)-\text{b}(\hat{\beta}x\vert d)
\end{align*} {#eq-general-gap}

where $\hat{\beta}$ is a coefficient computed from sample data and $\text{b}(z\vert w)$ is the slope from a regression of $z$ on $w$ and an intercept.

So equation @eq-general-gap gives $\text{Gap}^{1}$ when $\hat{\beta}$ is the slope coefficient of a regression of $y$ on $x$ using the data from group 1, and gives $\text{Gap}^{0}$ when $\hat{\beta}$ is the slope coefficient using data from group 0[^1].

[^1]: under the data generation process described by $y=\delta_{0}+\delta_{d}d+\delta_{1}X+\epsilon$

\begin{align*}
\text{Gap} & =\text{b}(y\vert d)-\text{b}(\hat{\beta}x\vert d)\\
 & =\frac{\text{cov}\left(d,y\right)}{\text{var}(d)}-\hat{\beta}\frac{\text{cov}\left(x,d\right)}{\text{var}(d)}\\
 & =\frac{\text{cov}\left(d,\beta_{0}+\beta_{d}d+\beta_{x}x\right)+\epsilon}{\text{var}(d)}-\hat{\beta}\frac{\text{cov}\left(d,x\right)}{\text{var}(d)}\\
 & =\beta_{d}+\beta_{x}\frac{\text{cov}\left(d,x\right)}{\text{var}(d)}-\hat{\beta}\frac{\text{cov}\left(d,x\right)}{\text{var}(d)}\\
 & \rightarrow\beta_{d},\;\text{when }\beta_{x}=\hat{\beta}
\end{align*} {#eq-specific-gap}

implying that all the gap expressions are equivalent, except for $\text{Gap}^{p}$. The difference arises because in the pooled regression used to compute $\text{Gap}^p$, $d$ is a omitted variable.

\begin{align*}
\text{Gap}^{p} & =\text{b}(y\vert d)-\text{b}(x\text{b}(y\vert x)\vert d)\\
 & =\frac{\text{cov}\left(d,y\right)}{\text{var}(d)}-\frac{\text{cov}\left(d,\left[\text{cov}\left(x,y\right)/\text{var}\left(x\right)\right]\right)}{\text{var}(d)}\\
 & =\frac{\text{cov}\left(d,y\right)}{\text{var}(d)}-\frac{\text{cov}\left(d,x\right)}{\text{var}(d)}\times\frac{\text{cov}\left(x,y\right)}{\text{var}(x)}\\
 & =\frac{1}{\text{var}(d)}\left(\text{cov}\left(d,y\right)-\frac{\text{cov}\left(d,x\right)\text{cov}\left(x,y\right)}{\text{var}(x)}\right)
\end{align*}

Compare this to the $\text{Gap}^{\text{OLS}}$, defining $\tilde{z}(w)$ as the component of $z$ that is orthogonal to $w$ in the population (so that $\tilde{z}(w)=z-w\text{b}(z\vert w)$, alternatively $\tilde{z}(w)=z-w\frac{\text{cov}(z,w)}{\text{var}(w)}$)[^2]

[^2]: Note: $\tilde{d}(x)=d-P(x)=d-x\frac{\text{cov}(x,d)}{\text{var}(x)}$, the linear propensity, and $\tilde{y}(x)=y-x\frac{\text{cov}(y,x)}{\text{var}(x)}$, so the coefficient $\delta_d$ is $\frac{\text{cov}(\tilde{d}(x),\tilde{y}(x))}{\text{var}(\tilde{d}(x))}$.

\begin{align*}
\text{Gap}^{\text{OLS}} & =\delta_{d}\\
 & =\frac{\text{cov}(\tilde{d}(x),\tilde{y}(x))}{\text{var}(\tilde{d}(x))};\;\text{per FWL}\\
 & =\frac{\text{cov}(d,\tilde{y}(x))}{\text{var}(\tilde{d}(x))}-\frac{\text{cov}(x,\tilde{y}(x))}{\text{var}(\tilde{d}(x))}\times\frac{\text{cov}(d,x)}{\text{var}(x)};\;\text{per definition of }\tilde{d}(x)\\
 & =\frac{\text{cov}(d,\tilde{y}(x))}{\text{var}(\tilde{d}(x))};\;\text{per definition of }\tilde{y}(x),\,\text{cov}(x,\tilde{y}(x))=0\\
 & =\frac{1}{\text{var}(\tilde{d}(x))}\left(\text{cov}(d,y)-\text{cov}(d,x)\frac{\text{cov}(x,y)}{\text{var}(x)}\right)
\end{align*}

and so $\text{Gap}^{p}=\frac{\text{var}(\tilde{d}(x))}{\text{var}(d)}\text{Gap}^{\text{OLS}}$

### Useful expressions

The law of total variance, or variance decomposition formula is

$$
\text{var}(X)=\mathbb{E}[\text{var}(X\vert Y)]+\text{var}(\mathbb{E}[X\vert Y])
$$ where the first term is the between-group variance and the second term is the within-group variance.

In our problem, the variance decomposition is

\begin{align*}
\text{var}(x) & =\left(1-\pi\right)\left(\text{var}(x\vert d=0)+\left(\mathbb{E}\left[x\vert d=0\right]-\mathbb{E}\left[x\right]\right)^{2}\right)\\
 & +\pi\left(\text{var}(x\vert d=1)+\left(\mathbb{E}\left[x\vert d=1\right]-\mathbb{E}\left[x\right]\right)^{2}\right)
\end{align*} but we have

$$
\mathbb{E}\left[x\right]=\mathbb{E}\left[\mathbb{E}\left[x\vert y\right]\right]=\pi\mathbb{E}\left[x\vert d=1\right] + (1-\pi)\mathbb{E}\left[x\vert d=0\right]
$$ substituting the RHS for $\mathbb{E}\left[x\right]$ we have

\begin{align*}
\text{var}(x) & =\left(1-\pi\right)\left(\text{var}(x\vert d=0)+\left(\mathbb{E}\left[x\vert d=1\right]-\mathbb{E}\left[x\vert d=0\right]\right)^{2}\pi^{2}\right)\\
 & +\pi\left(\text{var}(x\vert d=1)+\left(\mathbb{E}\left[x\vert d=1\right]-\mathbb{E}\left[x\vert d=0\right]\right)^{2}(1-\pi)^{2}\right)
\end{align*}

but we have $\mathbb{E}\left[x\vert d=1\right]-\mathbb{E}\left[x\vert d=0\right]=\text{cov}(x,d)/\text{var}(d)$, and $\text{var}(d)=\pi(1-\pi)$, so

\begin{align*}
\text{var}(x) & =\left(1-\pi\right)\text{var}(x\vert d=0)+\pi\text{var}(x\vert d=1)+\frac{\text{cov}(d,x)^{2}}{\text{var}(d)^{2}}\times\left[(1-\pi)\pi^{2}+\pi(1-\pi)^{2}\right]\\
 & =\left(1-\pi\right)\text{var}(x\vert d=0)+\pi\text{var}(x\vert d=1)+\frac{\text{cov}(d,x)^{2}}{\text{var}(d)^{2}}\times\left[(1-\pi)\text{var}(d)+\pi\text{var}(d)\right]\\
 & =\left(1-\pi\right)\text{var}(x\vert d=0)+\pi\text{var}(x\vert d=1)+\frac{\text{cov}(d,x)^{2}}{\text{var}(d)}
\end{align*}

by the same logic

$$
\text{cov}(x,y)=\left(1-\pi\right)\text{cov}(x,y\vert d=0)+\pi\text{cov}(x,y\vert d=1)+\frac{\text{cov}(d,x)\text{cov}(d,y)}{\text{var}(d)}
$$

Now writing $\text{Gap}^{\text{OLS}}=w_1\text{Gap}^1 + w_0\text{Gap}^0$ with $w_1+w_0=1$.

### Gap weights

Based on @eq-specific-gap we have

$$
\text{Gap}^1=\frac{\text{cov}\left(d,y\right)}{\text{var}(d)}-\frac{\text{cov}\left(x,d\right)}{\text{var}(d)}\times\frac{\text{cov}\left(x,y\vert d=1\right)}{\text{var}(x\vert d=1)}
$$

and

$$
\text{Gap}^0=\frac{\text{cov}\left(d,y\right)}{\text{var}(d)}-\frac{\text{cov}\left(x,d\right)}{\text{var}(d)}\times\frac{\text{cov}\left(x,y\vert d=0\right)}{\text{var}(x\vert d=0)}
$$
